A Triple Origin for Twin Blue Stragglers in Close Binaries

### 2019-03-04 response_to_referees.txt

For comparison, if we compute the critical orbital periods
corresponding to the timescales for a specific triple to undergo
either a 1+3 or 2+3 interaction, and if we set the 1+3 and 2+3
timescales equal to the typical duration expected for mass transfer,
which is of order 1 Myr (i.e., corresponding to the expected duration
of the RGB phase, which is longer than is expected for the AGB phase),
then we obtain critical orbital periods of, respectively, 8.1 x 108
and 2.4 x 109 days.  The critical outer tertiary orbital periods
corresponding to such short interaction times are much longer than the
maximum predicted outer orbital period of the hypothesized white dwarf
tertiary in our scenario, and fall outside of the current range of
orbital periods shown on the x-axis in figure 2.  Hence, we do not
expect the mass transfer process to be interrupted by either 1+3 and
2+3 interactions.  Rather than include these critical orbital periods
in figure 2, we have included a paragraph on this in the text,
indicated in bold-faced font.

2) The authors fix the amount of angular momentum lost per unit mass to eta=3. How did they estimate this number? Please, provide some additional information and references. Could the value of this number be the reason of the discrepancy between the theoretical expectations and the simulation results in Fig. 6?

SPZ/NL: Indeed, this is one of these magic parameters that have
received relatively little attention in the recent literature, even
though it's value is very important. We adopt the approach as
advocated in Pols et al 1991 and Portegies Zwart 1995 with further
support from the analysis of Pols 2007 and Krticka etal 2011. We do not
really know how much angular momentum is lost per unit mass in a
non-conservative binary system, but interestingly enough our analysis
in this paper appears to be consistent with these earlier findings.
We comment on this in the main text in the added sentence near Eq. 3.

@@

3) The authors should elaborate more on the reason why there is a large discrepancy in the plots shown in Fig. 6. What could be the main reasons? And why for a donor of 1.2 Msun the discrepancy is larger?

SPZ/NL: Let's first point out that the trends are rather
consistent. The 1.2 Msun donor's seem to be doing something similar as
the 1.4Msun donors. Except that the more massive donors follow the
required trend in order to be able to explain the observed system.
Then, the inner orbits of the 1.2MSun donors tend to shrink, whereas
the more massive donors drive the expansion of the inner orbit.

4) The authors adopt an initial inner binary of 0.9 Msun-1.1 Msun. I understand that the calculations may be long and complicated, but they should present results for an additional run where they change the masses of the inner binary, as they have already done in the case of the third companion. These would make stronger their claim that the inner binary would evolve towards a mass ratio of unity.

SPZ/NL: As we pointed out in the manuscript, we do not really
understand the subtleties of angular momentum transport in
non-conservatively evolving binaries. In triples, additional
complications are introduced by the third component. I fear that this
process require much more study, which we think, would be really
interesting to pursue. However, for this manuscript we prefer to focus
on the initial conditions that are most likely to reproduce the
observed twin blue straggler, and for that discussion we think that the
1.2Msun donor does not lead to final parameters that are consistent
with this observed system.

@@

5) Other interesting plots would be the evolution in time of the
masses of the three stars and their spins, and, more importantly of
the inner binary eccentricity.

SPZ/NL: The eccentricity of the inner binary remains small. It remains
below 0.02 for an inner orbital separation of 0.2au, and below 0.01
for the tighter inner orbits. We add a comment to this effect in the
paper.

For the evolution of the masses of the three stars we have figure 6.
to allow for a discussion on the initial mass-ratio of the inner
binary, we add a new figure (7), in which we present the mass ratio of
the inner binary as a function of the total mass of the inner binary
(both normalized). Here we demonstrate that the inner binary star
grows gradually to more equal masses, and that this effect is stronger
when the initial inner mass ratio is smaller. Interestingly, the 
initial mass-ratio of 0.9MSun/1.1MSun appears to be less favorable for
growing two equal mass stars, when we start with a lower
initial mass ratio, for example using 0.7MSun/1.1MSun. We may refine
our conclusion that maybe a lower initial mass ratio for the inner
binary is favored over an initially more equal mass binary. However,
the observational constraints are insufficient to bolster this
conclusion at this point.

@@

Important - "Dynamically, it is difficult to form a short-period binary composed of two collision products during a collisional interaction in a star cluster." I am not clear if by this statement the authors are excluding an exchange into a BS-[WD or MS] binary of another BS, creating 7782. This scenario has been posed previously. More broadly, with one stroke of this sentence, the authors are ruling out a host of other pathways to this binary. If that is their intent, this merits more justification. (To be clear, I don't think that they need to rule out all other scenarios for their scenario to be interesting.)

SPZ/NL:  This is of course a very valid point.  We do not intend to concretely rule out other formation channels, but rather to highlight that a purely dynamical formation channel likely has a low probability of occurring.  This is because, as explained in Leigh & Sills (2011), the energetics of a 1+2 interaction involving a BS and a pre-existing BS-MS or BS-WD system with a long orbital period are such that it is unlikely that the result will be a BS-BS binary with a short orbital period as is observed for WOCS 7782. And, as shown in Mathieu & Geller (2009), there are only two short-period BS binaries in NGC 188, such that the relevant 1+2 interaction time should be for a specific short-period BS binary, which increases the expected interaction time considerably such that it will exceed the expected BS lifetime.  To save on space, we have included a very brief mention of this in the draft.

before it leaves the asymptotic giant branch. - At this point in the paper, it is not evident why this needs to be AGB mass transfer and not RG mass transfer. That is, at this point it is not "according to this scenario"?

SPZ/NL: Our apologies, since we were not very clear about this.  The scenario only works for donors on the giant branch; not on the Hertzsprung-gap and not on
the AGB. We have now made that clear in fig. 3 by painting these parts of
the curve indicating the stellar radius orange.

@@

More importantly, what is the significance of these assumptions? Is the goal of this paper simply to show that the proposed scenario can happen, in which case my last question is not as important. But if the goal is more generality, than these assumptions do matter.

SPZ/NL: Our prime objective is to demonstrate that this scenario
works. We do not really desire to cover parameter space or indicate in
which part of parameter space this scenario works. This may be very
interesting, but for a future paper.

"We assume initial component masses of m1=1.1M⊙ and m2=0.9M⊙ for the inner binary components, and m3 = 1.4M⊙ for the outer tertiary." - Where did these assumptions for m2 and m3 come from? What is there motivation? And essentially the same question for exploring the range of 1.2⊙ < m3 < 1.4⊙ in Section 3.1. (As an aside, the way this is written, I first thought that these assumptions were being made for the entire paper, not just this section of the paper, which caused me some concern. This was not clarified/resolved (at least for m3) until Sections 3.1 and 4.
Perhaps the authors might want to make clearer here that here these assumptions are only for parameter delineation?)

SPZ/NL: These are good points, and yes this is effectively our choice
for parameter delineation that we expected to work best for a “proof
of concept” illustration via our simulations.  For instance, if the
initial mass ratio of the inner binary is too large, then it can occur
that not enough mass can be transferred to bring the inner binary
toward a mass ratio of unity. We added an explanation to the text to
express and motivate our rather explicit choice of parameters.

@@

and the amount of angular momentum lost per unit mass η ≃ 3. Again, this number comes out of the blue?

SPZ/NL: Indeed, this is one of these magic parameters that have
received relatively little attention in the recent literature, even
though it's value is very important. We adopt the approach as
advocated in Pols et al 1991 and Portegies Zwart 1995 with further
support coming from the analysis of Pols 2007 and Krticka etal 2011. We do not
really know how much angular momentum is lost per unit mass in a
non-conservative binary system, but interestingly enough our analysis
in this paper appears to be consistent with these earlier findings.
We comment on this in the main text in the added sentence near Eq. 3.

Tidal effects during mass tran[s]fer have probably circularized the orbit, although some slight

@@

SPZ/NL:  Fixed.

eccentricity due to turbulent motion in the outer layers of the donor star may have induced a small e ≪ 0.1 eccentricity - This has been the canonical assumption, but in fact observations don't support it (e.g., Mathieu & Geller 2009) and recent theoretical papers indicate why it might not happen. If the authors wish to stipulate this as yet another assumption of the paper, that is fine, but should be noted as such.

SPZ/NL: We have left out the remark about the low eccentricity of the outer orbit.

@@

250 R⊙[ ]for the 100 R⊙[ ]donor star

SPZ/NL:  Fixed.

If length is an issue, I think Figure 4 could be sacrificed. In any case, it should be zoomed in so that the system itself is more easily seen. (To be honest, I am totally perplexed by "The two companion stars are represented as black bullets (to the right).")

SPZ/NL: We removed figure 4. The referee is right that it does not really add much to the paper.  The "black bullets to the right" was a typo, left over after many iterations of refinement.

@@

SPZ/NL: Very interesting point. Thanks for pointing this out.
We have edited both of these bullet points as recommended.

Specific notes on the abstract:

Broadly speaking, the abstract is rather unusual in that it is not at all an abstract of the paper but rather a restatement of the predictions in the current Summary. As I wrote above, it is very good that the authors are making explicit predictions. But an abstract is intended to permit a reader to be informed about the entire essential content of a paper, and as such this abstract does not seem appropriately constructed.

SPZ/NL: We think that the abstract maybe a bit controversial in comparison
with other abstracts, but we think that it covers the content of the
paper and the presented work.

### 2019-03-25 BSS7782.tex
minor edits for submission
some update

We propose a formation mechanism for twin blue stragglers (BSs) in
compact binaries that involves mass transfer from an evolved outer
tertiary companion on to the inner binary via a circumbinary disk.  We
apply this scenario to the observed double BS binary WOCS 7782 in the
old open cluster NGC 188, and show that its observed properties
\textbf{can be} reproduced within the context of the proposed model.
Based on this model, we predict the following properties for twin BSs:
(1) For the outer tertiary orbit, the initial orbital period should
lie between 220 days $\lesssim$ P$_{\rm out}$ $\lesssim$ 1100 days,
assuming initial masses for the inner binary components of $m_{\rm 1}
= 1.1$ M$_{\odot}$ and $m_{\rm 2} = 0.7$ to $0.9$M$_{\odot}$ and an
outer tertiary mass of $m_{\rm 3} = 1.4$ M$_{\odot}$.  \textbf{(2)}
After Roche-lobe overflow, the outer star turns into a white dwarf
(WD) of mass 0.43 to 0.54\,\MSun. There is a correlation between the
mass of this WD and the outer orbital period: more massive WDs will be
on wider orbits.  (3) The rotational axes of both BSs will be aligned
with each other and the orbital plane of the outer tertiary WD. (4)
The BSs will have roughly equal masses, independent of their initial
masses (since the lower mass star accretes the most).  The dominant
accretor should, therefore, be enriched more effectively by the
accreted material.  As a result, one of the BSs will appear to be more
enriched by either He, C and O or by s-process elements, depending on
whether the donor started to overflow its Roche lobe on, respectively,
the red giant or {\bf early} asymptotic giant branch.  (5) Relative to
old dense clusters with high-velocity dispersions, twin BSs in close
binaries formed from the proposed mechanism should be more frequent in
the Galactic field and younger open clusters with ages $\lesssim$ 4-6
Gyr, since then the donor will have a radiative envelope. (6) the
orbit of the binary BS will have a small semi-major axis (typically
$\aplt 0.3$\,au) and be close to circular ($e \aplt 0.1$).

@@

N.W.C.L. acknowledges support from a Kalbfleisch Fellowship at the
American Museum of Natural History.  SPZ would like to thank Norm
Murray and CITA for their hospitality during my long-term visit.  This
work was supported by the Netherlands Research School for Astronomy
(NOVA). 
%%
In this work we use the matplotlib
\citep{2007CSE.....9...90H}, numpy
\citep{Oliphant2006ANumPy}, AMUSE
\citep{portegies_zwart_simon_2018_1443252}, SeBa
\citep{2012ascl.soft01003P}, Huayno \citep{2012NewA...17..711P}, MESA
\citep{2010ascl.soft10083P}, and GadGet2 \citep{2000ascl.soft03001S}
packages. The calculations ware performed using the LGM-II (NWO grant
\# 621.016.701) and the Dutch National Supercomputer at SURFSara
(grant \# 15520).

>>>>

We propose a formation mechanism for twin blue stragglers (BSs) in
compact binaries that involves mass transfer from an evolved outer
tertiary companion on to the inner binary via a circumbinary disk.  We
apply this scenario to the observed double BS-binary WOCS~7782 in the
old open cluster NGC~188, and show that its observed properties
\textbf{can be} reproduced within the context of the proposed model.
Based on this model, we predict the following properties for twin BSs:
(1) For the outer tertiary orbit, the initial orbital period should
lie between 220 days $\lesssim$ P$_{\rm out}$ $\lesssim$ 1100 days,
assuming initial masses for the inner binary components of $m_{\rm 1}
= 1.1$ M$_{\odot}$ and $m_{\rm 2} = 0.7$ to $0.9$M$_{\odot}$ and an
outer tertiary mass of $m_{\rm 3} = 1.4$ M$_{\odot}$.  \textbf{(2)}
After Roche-lobe overflow, the outer star turns into a white dwarf
(WD) of mass 0.43 to 0.54\,\MSun. There is a correlation between the
mass of this WD and the outer orbital period: more massive WDs will be
on wider orbits.  (3) The rotational axes of both BSs will be aligned
with each other and the orbital plane of the outer tertiary WD. (4)
The BSs will have roughly equal masses, independent of their initial
masses (since the lower mass star accretes the most).  The dominant
accretor should, therefore, be enriched more effectively by the
accreted material.  As a result, one of the BSs will appear to be more
enriched by either He, C and O or by s-process elements, depending on
whether the donor started to overflow its Roche lobe on, respectively,
the red giant or {\bf early} asymptotic giant branch.  (5) Relative to
old dense clusters with high-velocity dispersions, twin BSs in close
binaries formed from the proposed mechanism should be more frequent in
the Galactic field and younger open clusters with ages $\lesssim$ 4-6
Gyr, since then the donor will have a radiative envelope. (6) the
orbit of the binary BS will have a small semi-major axis (typically
$\aplt 0.3$\,au) and be close to circular ($e \aplt 0.1$).

@@

The authors kindly thank two anonymous reviewers for their considerable insight and 
suggestions for improvement.  N.W.C.L. acknowledges support from a Kalbfleisch Fellowship at the
American Museum of Natural History.  SPZ would like to thank Norm
Murray and CITA for their hospitality during my long-term visit.  This
work was supported by the Netherlands Research School for Astronomy
(NOVA). 
%%
In this work we use the matplotlib
\citep{2007CSE.....9...90H}, numpy
\citep{Oliphant2006ANumPy}, AMUSE
\citep{portegies_zwart_simon_2018_1443252}, SeBa
\citep{2012ascl.soft01003P}, Huayno \citep{2012NewA...17..711P}, MESA
\citep{2010ascl.soft10083P}, and GadGet2 \citep{2000ascl.soft03001S}
packages. The calculations ware performed using the LGM-II (NWO grant
\# 621.016.701) and the Dutch National Supercomputer at SURFSara
(grant \# 15520).
